---
title: "Data Preprocessing"
author: "J. Lecumberri, B. Fité & OBustos"
date: "2024-04-09"
output:
  html_document:
    toc: yes
    number_sections: yes
    toc_depth: 5
    toc_float:
      collapsed: yes
      smooth_scroll: yes
  pdf_document:
    toc: yes
    toc_depth: '5'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Libraries loading

```{r, warning=FALSE}
library(ggplot2)
library(stringr)
library(readxl)
library(data.table)
library(openxlsx)
library(knitr)
library(kableExtra)
library(RColorBrewer)
library(scales)
library(dplyr)
library(gridExtra)
library(plotly)
library(car)
```

# Data retrieval

Data of the project will be stored in variable `data`.

```{r}
cd = getwd()
parent_dir <- dirname(cd)
file_path_data = file.path(parent_dir, "data")

data = read.table(file.path(file_path_data, "heart_data.csv"), header = T, sep = ",")
```

## Initial data exploration

Our `data` is initially comprised of 70000 observations and 14 variables

```{r}
dim(data)
cat("Variable names: \n", colnames(data))
```

These variables are:

* **index** and **id**: subject information. Only id will be kept.
* **age**: age in days of each subject.
* **gender**: binary variable (0 and 1) with gender information.
* **height**: data in cm.
* **weight**: data in kg.
* **ap_hi**: systolic blood pressure reading.
* **lo_hi**: diastolic blood pressure reading.
* **cholesterol**: total cholesterol level on scale 0-5. Each unit denotes decrease by 20 mg/dL.
* **gluc**: glucose level read as mmol/l.
* **smoke**: wether subject smokes (1) or not (0).
* **alco**: whether subject drinks alcohol (1) or not (0).
* **active**: whether subject is active (1) or not (0).
* **cardio**: whether subject suffers from cardiovascular disease (1) or not (0).


# Data processing

Before any analysis it is important to perform the necessary modifications, additions and filtering steps in order to correct units, missing values, add informative data or remove outliers.

## Initial variable modifications and additions

First of all we will correct `age` variable and put it in years units:

```{r}
data$age = round(data$age / 365, 2) 
```

Addition of BMI variable, computed as follows:

$$BMI = \frac{weight_{kg}}{height_{m}^2}$$
```{r}
data$BMI = round(data$weight / (data$height / 100)**2, 2) 
```

Removal of `id` and `index` variables as row id already covers for this information:

```{r}
data = data[,c(-1,-2)] 
```

## Filtering: removal of outliers

By inspecting how the different variables are:

### Height - Weight

```{r}
boxplot(data$height, main = "HEIGHT")
boxplot(data$weight, main = "WEIGHT")

ggplot(data, aes(x = height, y = weight)) +
  geom_point() +
  labs(x = "height", y = "weight") +
  theme_minimal()
```

We can see how in `height` and `weight` have a lot of clears outliers and clear data errors:

```{r}
cat("Maximum value of height", max(data$height),"\n")
cat("Maximum value of weight", max(data$weight),"\n")
cat("Minumum value of height", min(data$height),"\n")
cat("Minumum value of weight", min(data$weight),"\n")
```
Filtering applied:

```{r}
data <- data[data$height >= 130 & data$height <= 230, ] 
data <- data[data$weight >= 45,]
```

Once filtered is looks as follows:

```{r}
ggplot(data, aes(x = height, y = weight)) +
  geom_point() +
  labs(x = "height", y = "weight") +
  theme_minimal()
```


### Systolic and diastolic pressures

First of all let's check how these two variables are distributed:

```{r}
ggplot(data, aes(x = ap_hi, y = ap_lo)) +
  geom_point() +
  labs(x = "ap_hi", y = "ap_lo") +
  theme_minimal()
```

We can observe clear outliers:

```{r}
cat("Maximum value of systolic P", max(data$ap_hi),"\n")
cat("Maximum value of diastolic P", max(data$ap_lo),"\n")
cat("Minumum value of systolic P", min(data$ap_hi),"\n")
cat("Minumum value of diastolic P", min(data$ap_lo),"\n")
```
So we will filter data by sources found for normal values of systolic and diastolic pressure

```{r}
data <- data[data$ap_hi <= 200 & data$ap_lo <= 140, ]
data <- data[data$ap_hi >= 80 & data$ap_lo >= 50, ]

ggplot(data, aes(x = ap_hi, y = ap_lo)) +
  geom_point() +
  labs(x = "ap_hi", y = "ap_lo") +
  theme_minimal()
```

### Gender detection

We have a binary result for gender: 1 and 2, but we need to characterize it correctly by male and female.

```{r}
weights_1 = data[data$gender == 1,]$weight
weights_2 = data[data$gender == 2,]$weight

cat("Mean weight value for gender = 1",mean(weights_1),"\n")
cat("Mean weight value for gender = 2",mean(weights_2),"\n\n")

heights_1 = data[data$gender == 1,]$height
heights_2 = data[data$gender == 2,]$height

cat("Mean height value for gender = 1",mean(heights_1),"\n")
cat("Mean height value for gender = 2",mean(heights_2))
```


So knowing that males, even more in a dataset of this size, have in avergare higher height and weight:

```{r}
data$gender <- ifelse(data$gender == 1, "F", "M")
data$gender <- as.factor(data$gender)
```


```{r}
ggplot(data, aes(x = weight, y = height, color = gender)) +
  geom_point() +
  labs(x = "Weight", y = "Height", color = "Gender") +
  theme_minimal()
```

```{r}
ggplot(data, aes(x = ap_hi, y = ap_lo, color = gender)) +
  geom_point() +
  labs(x = "Systolic pressure", y = "Diastolic pressure", color = "Gender") +
  theme_minimal()
```


```{r}
par(mfrow=c(1,3))
plot(density(data$age), main = "Age")
plot(density(data$height), main = "Height")
plot(density(data$weight), main = "Weight")
plot(density(data$ap_hi), main = "Ap Hi")
plot(density(data$ap_lo), main = "Ap Lo")
plot(density(data$BMI), main = "BMI")
```

We have normal apparent normal distribution across all these continuous variables. 

### Factorization of discrete variables

Now for the finale variables, those that are discrete over certain ranges of values, specified in the variables descriptions, we can study their representations and factor them.

```{r}
table(data$cholesterol)
table(data$gluc)
table(data$smoke)
table(data$alco)
table(data$active)
table(data$cardio)

data$cholesterol <- as.factor(data$cholesterol)
data$gluc <- as.factor(data$gluc)
data$smoke <- as.factor(data$smoke)
data$alco <- as.factor(data$alco)
data$cardio <- as.factor(data$cardio)
```

```{r}
p1 <- ggplot(data, aes(x = cholesterol)) +
  geom_bar() +
  labs(title = "Cholesterol Levels",
       x = "Cholesterol Levels",
       y = "Count") +
  theme(plot.title = element_text(size = 10)) # Adjust size here

p2 <- ggplot(data, aes(x = gluc)) +
  geom_bar() +
  labs(title = "Glucose Levels",
       x = "Glucose Levels",
       y = "Count") +
  theme(plot.title = element_text(size = 10)) # Adjust size here

p3 <- ggplot(data, aes(x = smoke)) +
  geom_bar() +
  labs(title = "Smoke Status",
       x = "Smoke Levels",
       y = "Count") +
  theme(plot.title = element_text(size = 10)) # Adjust size here

p4 <- ggplot(data, aes(x = alco)) +
  geom_bar() +
  labs(title = "Alcohol Status",
       x = "Alcohol Levels",
       y = "Count") +
  theme(plot.title = element_text(size = 10)) # Adjust size here

p5 <- ggplot(data, aes(x = active)) +
  geom_bar() +
  labs(title = "Active Status",
       x = "Active Levels",
       y = "Count") +
  theme(plot.title = element_text(size = 10)) # Adjust size here

p6 <- ggplot(data, aes(x = cardio)) +
  geom_bar() +
  labs(title = "Cardio Output Levels",
       x = "Cardio Levels",
       y = "Count") +
  theme(plot.title = element_text(size = 10)) # Adjust size here

grid.arrange(p1, p2, p3, p4, p5, p6, nrow = 2, ncol = 3)
```

